\begin{thebibliography}{}

\bibitem{A:1977}  
{\sc Ames, W. F. (1977).} \emph{Numerical Methods for Partial
  Differential Equations.} Academic Press.

\bibitem{AH:2002}  
{\sc Angelini, F., \& S. Herzel (2002).} Consistent Initial Curves
for Interest Rate Models, \emph{Journal of Derivatives} 9(4), 8--18.

\bibitem{AH:2005}
{\sc Angelini, F., \& S. Herzel (2005).} Consistent Calibration of
HJM Models to Cap Implied Volatilities, \emph{Journal of Futures Markets} 25, 1093-1120.

\bibitem{BIS:2005}
BIS (2005). Zero-coupon yield curves: Technical documentation, \emph{BIS Papers}
25. Bank for International Settlements, Basle. 

\bibitem{B:2004}		
{\sc Bj\"ork, T. (2004).} \emph{Arbitrage Theory in Continuous Time}, Second Edition, Oxford.

\bibitem{B:2003}
{\sc Bj\"ork, T. (2003).} On the Geometry of Interest Rate Models, en:
Carmona, R.A., Cinlar, E., Ekeland, I., Jouini, E., Scheinkman, J.A., Touzi, N.,
\emph{Paris-Princeton Lectures on Mathematical Finance 2003}, Lecture Notes in
Mathematics, 1847, Springer.  

\bibitem{BL:2002}
{\sc Bj\"ork, T., \& C. Landen (2002).} On the construction of finite
dimensional realizations for nonlinear forward rate models,
\emph{Finance and Stochastics} 6(3), 303-331. 

\bibitem{BS:2001}
{\sc Bj\"ork, T., \& L. Svensson (2001).} On the existence of finite
dimensional realizations for nonlinear forward rate models,
\emph{Mathematical Finance} 11(2), 205-243. 

\bibitem{B:2001}
{\sc Bj\"ork, T. (2001).} A Geometric View of Interest Rate Theory, en:
Jouini, E., Cvitanic, J. \&  M. Musiela, \emph{Option Pricing, Interest
Rates and Risk Management}, Cambridge University Press.  

\bibitem{BG:1999}
{\sc Bj\"ork, T., \& A. Gombani (1999).} Minimal Realizations of interest rate
models. \emph{Finance and Stochastics} 3(4), 413-432.

\bibitem{BC:1999}
{\sc Bj\"ork, T., \& B.J. Christensen (1999).} Interest Rate Dynamics
and Consistent Forward Rate Curves, \emph{Mathematical Finance} 9,
323-348.  

\bibitem{B:1997}
{\sc Bj\"ork, T. (1997).} Interest Rate Theory, en: Runggaldire, W., \emph{Financial Mathematics}, Lecture Notes in Mathematics, 1656, Springer-Verlag. 

\bibitem{BK:1991}
  {\sc Black, F., \& P. Karasinski (1991).} Bond and Option Pricing when
Short Rates are Lognormal, \emph{Financial Analysts Journal} 4, 52-59.
  127--155.  

\bibitem{B:1976}
{\sc Black, F.}  The Pricing of Commodity Contracts, \emph{Journal of
  Financial Economics} 3, 167--179.

\bibitem{BM:2006}
{\sc Brigo, D. \& F. Mercurio (2006).} \emph{Interest Rate Models:
  Theory and Practice.} Cambridge University Press. 

\bibitem{BCo:2005}
{\sc Buraschi, A., \& F. Corielli (2005).} Risk Management
Implications of Time-Inconsistency: Model Updating and Recalibration
of No-Arbitrage Models. \emph{Journal of Banking \& Finance} 29, 2883-2907.

\bibitem{C:2004}
{\sc Cairns, A.~J.~G (2004).} \emph{Interest Rate Models. An
  Introduction} Princeton University Press. 

\bibitem{Ch:1996}
{\sc Cheyette, O. (1996).} Markov Representation of the
Heath-Jarrow-Morton Model. \emph{BARRA Inc.}. Working Paper.

\bibitem{PZ:1992}
{\sc Da Prato, G. \& J. Zabczyk  (1992).} \emph{Stochastic Equations in Infinite
  Dimensions.} Cambridge University Press.

\bibitem{R:2004}
{\sc De Rossi, G. (2004).} Kalman Filtering of Consistent Forward
Rate Curves: a Tool to Estimate and Model Dynamically the Term
Structure, \emph{Journal of Empirical Finance} 11, 277--308.  

\bibitem{D:2010}
{\sc Daglish, T. (2010).}
Lattice Methods for No-Arbitrage Pricing of Interest Rate Securities,
\emph{Journal of Derivatives} 18(2), 7--19.  

\bibitem{DHW:1995}
{\sc Dewynne, J., Howison S. \& P. Wilmott (1995).}
\emph{The Mathematics of Financial Derivatives}, Cambridge University
Press.    

\bibitem{EKO:1990}
{\sc Eschenauer, H., Koski J. \& A. Osyczka (1990).} 
\emph{Multicriteria Design Optimization}, Springer.  


\bibitem{FNN:2011}
{\sc Falc\'o A., Navarro Ll. \& J. Nave. (2011).} On the Calibration of a Gaussian
Heath-Jarrow-Morton model using Consistent Forward Rate
Curves. \emph{Quantitative Finance} 11(4), 495--504. 

\bibitem{FNN:2009}
{\sc Falc\'o A., Navarro Ll. \& J. Nave (2009).} The Hull-White Model
and Multiobjective Calibration with Consistent Curves: Empirical
Evidence. \emph{RACSAM Rev. R. Acad. Cien. Serie A. Mat.} 103(2), 235--249.  

\bibitem{FNN:2008}
{\sc Falc\'o A., Navarro Ll., \& J. Nave (2008)}. A multiobjective approach using
consistent rate curves to the calibration of a Gaussian
Heath-Jarrow-Morton model. \emph{WP-AD Series} 2008--09, IVIE. 

\bibitem{Fil:2009}
{\sc Filipovic D., (2009).} \emph{Term-Structure Models}, Springer. 

\bibitem{Fil:2001}
{\sc Filipovic, D., (2001).} \emph{Consistency Problems for
Heath-Jarrow-Morton Interest Rate Models}, Lecture Notes in
Mathematics, 1760, Springer-Verlag. 

\bibitem{Fil:1999}
{\sc Filipovic D., (1999).} A Note on the Nelson and Siegel
Family. \emph{Mathematical Finance} 9(4), 349--359.  

\bibitem{HaKreps:1978}
{\sc Harrison J.M. \& D Kreps (1978).} Martingales and Arbitrage in
Multiperiod Securities Markets. \emph{Journal of Economic Theory} 20,
381--408.
\bibitem{HJM:1992}
{\sc Heath, D.~C., R. Jarrow \& A. Morton (1992).} Bond Pricing and the
Term Structure of Interest Rates: A New Methodology for Contingent
Claims Valuation, \emph{Econometrica} 60(1), 77--105.  

\bibitem{HL:1986}
{\sc Ho, T.S.Y, \& S.B. Lee (1986).} Term Structure Movements and the
Pricing of Interest Rate Contingent Claims. \emph{Journal of Financial
  and Quantitative Analysis} 411011-1029.

\bibitem{HW:1990}
{\sc Hull, J., \& A. White (1990).} Pricing Interest Rate
Derivatives Securities. \emph{The Review of Financial Studies}
3(4), 573--592.

\bibitem{HW:1993}
{\sc Hull, J., \& A. White (1993).} Single-factor interest rate
models and the valuation of interest rate derivative
securities. \emph{Journal of Financial and Quantitative Analysis}
3(4), 573--592. 

\bibitem{HW:1994}
{\sc Hull, J., \& A. White (1994).} Numerical Procedures for
Implementing Term Structure Models, I: Single-Factor Models. \emph{The
Journal of Derivatives} 2, 7--16. 

\bibitem{IW:1981}
{\sc Ikeda, N., \& S. Watanabe (1981).} \emph{Stochastic Differential
Equations and Diffusion Processes}, North-Holland.

\bibitem{J:1991}
{\sc Jamshidian, F. (1991).} Forward Induction and Construction of
Yield Curve Difussion Models. \emph{The Journal of Fixed Income} 1(1),
62--74.

\bibitem{KW:2005}
{\sc Kim, I.~Y., \& O.L. de Weck (2005).} Adaptive weighted-sum method
for bi-objective optimization: Pareto front
generation. \emph{Struct. Multidisc. Optim.} 29, 149--158.    

\bibitem{KP:1999}
{\sc Kloeden, P.~E., \& E. Platen (1999).} \emph{Numerical solution of
  stochastic differential equations}, Springer.

\bibitem{M:2005}
{\sc Marler, T., (2005).} A study of multi-objective optimization
methods for engineering applications, Ph.D. Dissertation,
\emph{University of Iowa}, Iowa City, IA. 

\bibitem{MM:2000}
{\sc Mercurio, F., \& J.M. Moraleda (2000).} An analytically tractable interest
rate model with humped volatility. \emph{European Journal of Operational
  Research} 120, 205--214.

\bibitem{Mu:1993}
{\sc Musiela, M., (1993)}. Stochastic PDEs and term structure models.
Working Paper, J. Intern. Finance,  IGR-AFFI, La Baule.

\bibitem{MR:2004}
{\sc Musiela, M., \& M. Rutkowski (2004)}. \emph{Martingale methods in
financial modelling}, Springer. 

\bibitem{N:2003}
{\sc Navarro, Ll., (2003).} M\'etodos Num\'ericos para la Valoraci\'on
de Derivados sobre Tipos de Inter\'es , Master Thesis Dissertation,
\emph{University of Valencia}, Valencia.  

\bibitem{NS:1987} 
{\sc Nelson, C. R., \& A.~F.~Siegel (1987).} Parsimonious Modelling of Yield
Curves. \emph{Journal of Business} 60(4), 473--489.

\bibitem{O:1998}
{\sc \O ksendal, B. (1998).} \emph{Stochastic Differential Equations}. Springer

\bibitem{P:1906}
{\sc Pareto, V. (1906).} Manuale di Economia Politica. \emph{Societa
  Editrice Libraria}.

\bibitem{R:1987}
{\sc Rao, S. S. (1987).} Game Theory Approach for Multiobjective
Structural Optimization. \emph{Computers and Structures} 25(1), 119--127.

\bibitem{RC:1999}
{\sc Ritchken, P., \& I. Chuang (1999).} Interest rate option pricing with
volatility humps. \emph{Review of Derivatives Research} 3, 237--262.

\bibitem{RS:1992}
{\sc Ritchken, P. and L. Sankarasubramanian (1992).} On Markovian
Representations of the Term Structure, Working Paper.  

\bibitem{S:1985}
{\sc Smith, G.~D.~(1985).} \emph{Numerical Solution of Partial Differential
Equations}, Oxford University Press

\end{thebibliography}
